Starcode: sequence clustering based on all-pairs search

### 2014-11-07 doc/document.tex
Working on document.tex (preparing for merge conflicts).

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. In many
of those applications, sequencing errors need to be corrected.
This is particularly important when using random DNA barcodes to
trace and quantify transcripts or lineages, since spurious barcodes
introduce inexistent entities. The reference population of random
barcodes is unknown and in this case, error correction amounts
to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computational complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to
determine which pairs of sequences lie within a given Levenshtein
distance. For error correction or redundancy reduction purposes,
matched pairs are then merged into clusters represented by a
canonical sequence. The effiency of starcode is attributable to
the poucet search, a novel implementation of the Needleman-Wunsch
algorithm performed on the nodes of a trie. On the task of matching
random barcodes, starcode outperforms sequence clustering algorithms
in both speed and precision.

@@

Sequencing technologies never achieve perfect precision. For instance,
the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
consisting of substitutions \citep{pmid18660515, pmid21576222}
and the PacBio platform has a 15\% error rate consisting of insertions
and deletions \citep{pmid19023044}. The enormous throughput of such technologies
has recently created additional needs to develop efficient error
correction algorithms.

@@

Here we describe a sequence clustering algorithm focused on error
correction. The algorithm is called ``starcode'' in reference to 
clusters of random barcodes, which typically have a star shape.
Starcode is based on all-pairs search, \textit{i.e.} all the pairs
of sequences below a given Levenshtein distance are identified during
the graph construction phase. Matching is carried out by lossless
filtration, followed by an exhaustive search on the branches of a
prefix trie.  The novelty of the algorithm is the poucet strategy,
which uses the redundancy of alphabetically sorted sequences to avoid
unnecessary recomputations and gain speed.

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. File formats compatible
with starcode are: raw sequence, raw sequence with count, FASTA or
FASTQ (in which case starcode ignores the quality). Starcode has two
output formats: cluster and non-redundant. In cluster mode, starcode
returns detailed information of the clustering results, i.e. canonical
sequences, cluster sizes and the complete list of their constituent
sequences. In non-redundant mode, only the canonical sequences are
printed. The second mode is especially useful to filter out redundant
sequences from input files. By default, clustering is performed under
the assumption that divergence occurs from experimental errors
(sequencing errors, PCR mutations \textit{etc}.), a more general
algorithm is also available for other clustering problems (an example
of which is given in section \ref{section:motifs}).

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets (Figure \ref{fig:benchmark}a). We also observed that
slidesort found 5-10\% less 3-matches than starcode on all the
datasets (Figure \ref{fig:benchmark}b). We were surprised by 
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons are feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc).

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show linear and sub-linear trends,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this result, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Note that the comparison with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering (which is by far the most memory
demanding structure). The memory usage of starcode is the highest for
clusters of size 1000, but it decreases and becomes lower than the
memory usage of seed as the size of the cluster increases.

In conclusion, starcode was the only tool to achieve perfect precision
on these datasets at a price of increased memory
footprint. Considering the exactness of the output, starcode maintains
a competitive performance in terms of running time.

@@

\section*{Acknowledgement}
We would like to thank Maria Chatzou for her precious feedback on the
preliminary version of this manuscript.

>>>>

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. In many
of those applications, sequencing errors need to be corrected.
This is particularly important when sequencing reads from an
unknwon reference such as random DNA barcodes.
In this case, error correction can be
done by performing a pairwise comparison of all the barcodes,
which is computationally complex problem.

\section{Results:}
Here we address this challenge and describe an exact algorithm to
determine which pairs of sequences lie within a given Levenshtein
distance. For error correction or redundancy reduction purposes,
matched pairs are then merged into clusters of similar sequences.
The effiency of starcode is attributable to
the poucet search, a novel implementation of the Needleman-Wunsch
algorithm performed on the nodes of a trie. On the task of matching
random barcodes, starcode outperforms sequence clustering algorithms
in both speed and precision.

@@

All sequencing technologies have a certain degree of imprecision.
For instance, the Illumina platform \citep{pmid16056220} has a 1-2\%
error rate consisting of substitutions \citep{pmid18660515, pmid21576222}
and the PacBio platform has a 15\% error rate consisting of insertions
and deletions \citep{pmid19023044}. The enormous throughput of such
technologies has recently created additional needs for developing
efficient error correction algorithms.

@@

Here we describe a sequence clustering algorithm called ``starcode''
in reference to clusters of random barcodes, which typically have a star
shape.  Starcode is based on all-pairs search, \textit{i.e.} all the pairs
of sequences below a given Levenshtein distance are identified during
the graph construction phase. Matching is carried out by lossless
filtration, followed by an exhaustive search on the branches of a
prefix trie.  The novelty of the algorithm is the poucet strategy,
which uses the redundancy of alphabetically sorted sequences to avoid
unnecessary recomputations and gain speed.

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. File formats compatible
with starcode are raw sequence, raw sequence with count, FASTA or
FASTQ (in which case starcode ignores the quality). Starcode either
returns detailed information of the clustering results, i.e. canonical
sequences, cluster sizes and the complete list of their constituent
sequences. Alternatively, only the canonical sequences are
printed, which is useful to filter out redundant
sequences from input files. By default, clustering is performed under
the assumption that divergence occurs from experimental errors
(sequencing errors, PCR mutations \textit{etc}.) and a more general
algorithm is also available for other clustering problems (an example
of which is given in section \ref{section:motifs}).

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets (Figure \ref{fig:benchmark}a). We also observed that
slidesort found 5-10\% less 3-matches than starcode on all the
datasets (Figure \ref{fig:benchmark}b). We were surprised by 
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons are feasible
(more information is available on the starcode repository the, see
http://github.com/gui11aume/starcode/misc).

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. The running
time of slidesort and starcode show linear and sub-linear trends,
respectively. Seed and cd-hit run approximately in constant time
regardless of the cluster size. In spite of this result, the
performance of starcode remains competitive, even for clusters of 1
million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c. The smallest memory footprint is achieved by
slidesort and cd-hit, with a maximum difference of an order of
magnitude with respect to the other tools. Note that the comparison with
slidesort is not completely fair since it does not hold in memory the
full graph necessary for clustering. The memory usage of starcode is
the highest for clusters of size 1000, but it decreases and becomes
lower than the memory usage of seed as the size of the cluster increases.
In conclusion, starcode was the only tool to achieve perfect precision
on these datasets at a price of increased memory
footprint. Considering the exactness of the output, starcode maintains
a competitive performance in terms of running time.

@@

\section*{Acknowledgement}
We would like to thank Maria Chatzou for her precious feedback on the
preliminary version of this manuscript and Heng-Chang Chen for
performing the Drosophila TRIP experiments used for benchmarking.
